---
title: "STAT-401 Group Project"
author: Chao-Wei Tu, Eugene Domrachev, Sean Kirker, Bintou Sako, Anna Tufail, Kyle
  Weston
output:
  html_document:
    df_print: paged
---

## Import Library / Data
```{r}
library(ggplot2)
library(data.table)
library(ggplot2)
library(gganimate)
library(transformr)
library(dplyr)
library(tidyverse)
library(scales)

#Cases from JLU
confirmed_cases_US <- "https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"
confirmed_cases_US <- read.csv(confirmed_cases_US)
#Deaths from JLU
deaths_US <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"
deaths_US <- read.csv(deaths_US)

owid_vaccines <- "https://raw.githubusercontent.com/owid/covid-19-data/13ee5b4126177156fba70fada48f7905da60e9c4/public/data/vaccinations/us_state_vaccinations.csv"
owid_vaccines <- read.csv(owid_vaccines)


```

## Data Cleansing
```{r}

vaccines_time = owid_vaccines %>% select("date"| "location" | "people_vaccinated")
head(vaccines_time)


cases = confirmed_cases_US %>% select("Province_State" | starts_with("X"))
deaths = deaths_US %>% select("Province_State" | starts_with("X"))

cases = cases %>% group_by(Province_State) %>% summarise(across(starts_with("X"), sum))
deaths = deaths %>% group_by(Province_State) %>% summarise(across(starts_with("X"), sum))

setDT(cases)
cases = cases %>% melt(cases, id=c("Province_State"), measure=patterns("^X"), value.name="Cases", variable.name="Date")

setDT(deaths)
deaths = deaths %>% melt(deaths, id=c("Province_State"), measure=patterns("^X"), value.name="Deaths", variable.name="Date")

cases$Date = as.Date(cases$Date, format="X%m.%d.%y")
deaths$Date = as.Date(deaths$Date, format="X%m.%d.%y")
covid <- merge(cases, deaths, by=c("Province_State", "Date"))

state = "Maryland"
state_covid = covid %>% filter(Province_State == state)

#reshape data from 'wide' to 'long' for princess ggplot 
state_covid = state_covid %>% pivot_longer(cols= c(Deaths, Cases), names_to = "Type")
head(state_covid)

vaccines_time$date = as.Date(vaccines_time$date)   
state_vax = vaccines_time %>% filter(location == state)
head(state_vax)

```

## Maryland Covid Stats
```{r}
theme_set(theme_gray())


death_plot <- ggplot(data = state_covid, aes(x = Date,y = value , fill = Type)) +
  labs(title = "Covid Cases and Deaths in Maryland"  ) +
  geom_area(alpha = .65)  +
  scale_y_continuous(name = "Number of Cases/Deaths", labels = comma) +
  scale_fill_manual(values = c('orange','red'))
death_plot
death_plot + transition_reveal(Date)
```



```{r}

vax_plot <- ggplot(data = state_vax, aes(x = date, y = people_vaccinated)) +
  labs(title = "Covid Vaccines in Maryland"  ) +
    geom_area( fill="darkblue", alpha = 0.67) 
    
vax_plot
vax_plot + transition_reveal(date) 

```

# Hypothesis Test on Proportion of Cases vs Death
```{r}
#Let's do a hypothesis test of a population proportion:
#First get the data from the latest day as of (4-16)
deaths_4_16 <- state_covid[[901,4]]
cases_4_16 <- state_covid[[902,4]]
deaths_4_16
cases_4_16
```

We test the null hypothesis for the population parameter $p$ which represents the proportion of deaths out of cases.
Using the assumption that death represents 1% of cases, we have $H_0:p=0.01$ vs $H_a:p\neq0.01$. 
This means we will use a two-tailed test.
$ n\cdot p_0 = 433359 \cdot 0.01 = 4333.59 > 10$ and $n \cdot q_0 = 433359 \cdot .99 = 429025.4 > 10$ So we can use a large sample Z-test.

```{r}
#calculate the Z-stat
calc_z_stat <- function(p_hat, p_0, n){
  return((p_hat-p_0)/sqrt(((p_0*(1-p_0))/n)))
}

p_hat = deaths_4_16/cases_4_16
p_hat #This is an unbiased estimator for the proportion of deaths
z_stat <- calc_z_stat(p_hat, 0.01, cases_4_16)
z_stat
2*(pnorm(abs(z_stat), lower.tail = FALSE)) 


```
However, we reject the null hypothesis a little too conclusively. Lets find a suitable sample size for $a = 0.01$ and $b = 0.05$ 
with $p' = 0.019$ (close to the $\hat{p}$ we calculated earlier)

```{r}
calc_samp_size <- function(alpha, p_0, beta, p_prime) {
  z_a <- qnorm(alpha, lower.tail = FALSE)
  z_b <- qnorm(beta, lower.tail = FALSE)
  return(( (z_a*sqrt(p_0*(1-p_0)) + z_b*sqrt(p_prime*(1-p_prime)))/(p_prime - p_0))^2 )
}

n <- ceiling(calc_samp_size(0.01, 0.01, 0.05, 0.019))
n
```
So we need a sample size of 2568.
Assuming that covid deaths are contained within cases, lets sample from our current sample at the appropriate sample size.
```{r}

#make a vector to sample from
#433359 - 8528 = 424831
x <- c(replicate(424831,0),replicate(8528,1))

mean(x) #matches p_hat

#sample the sample
set.seed(0)
samp <- sample(x,n)
mean(samp)

#calc z test (again) this time with an upper tailed test
z_stat <- calc_z_stat(mean(samp), 0.01, n)
z_stat
pnorm(z_stat, lower.tail = FALSE)

```
Once again we reject the null hypothesis.


# Hypothesis on Linear Regression Between Vaccinces and Deaths
We are interested in seeing if there is a negative relationship between the number of vaccinated people versus the number of death.

Here are some functions that might be helpful for the linear regressions later.
```{r}
sxx <- function(x){
  x_bar = mean(x)
  return(Reduce('+', (lapply(x, function(e)(e - x_bar)^2))))
}

syy <- function(y){
  y_bar = mean(y)
  return(Reduce('+', (lapply(y, function(e)(e - y_bar)^2))))
}

sxy <- function(x, y){
  x = as.numeric(x)
  x_bar = mean(x)
  y_bar = mean(y)
  
  s = 0
  for(i in 1:length(x)){
    s = s + (x[i] - x_bar)*(y[i] - y_bar)
  }
  return(s)
}
```


Unfortunately, there are some missing data for the number of doses administered to people in some days. We will first find a linear model of the number of vaccine administered as a function of time, then we use this model to fill in the missing data.
```{r}
# we omit rows that has NA and use the rest to find our linear model
tt_vac <- na.omit(vaccines_time %>% filter(location == "United States"))
x_samp = as.numeric(tt_vac$date)
y_samp = tt_vac$people_vaccinated

x_bar = mean(x_samp)
y_bar = mean(y_samp)

# using the functions earlier, we can find an estimate of the parameters of the linear model
beta_hat <- sxy(x_samp, y_samp)/sxx(x_samp)
alpha_hat <- y_bar - beta_hat * x_bar
y_hat <- function(x){beta_hat*x + alpha_hat}

# we then find the day that has the missing data, use the model to predict how many doses were administered that day. Put zero if the predicted model gives a negative value.
tt_vac <- vaccines_time %>% filter(location == "United States")
x <- as.numeric(tt_vac$date)
y <-tt_vac$people_vaccinated
for(i in 1:length(x)){
  if(is.na(y[i])){
    y[i] <- if(y_hat(x[i]) > 0) y_hat(x[i]) else 0
  }
}

tt_vac$people_vaccinated <- y

# plot
plot(x_samp, y_hat(x_samp), type='l', main = "United States", xlab = "date", ylab = "people vaccinated", xaxt='n',xlim = c(min(x),max(x)), ylim = c(min(y),max(y)))
axis(1, at=x[seq(1, length(x), by=length(x)/6)], labels=tt_vac$date[seq(1, length(x), by=length(x)/6)])
points(x, y)

```
From the plot, we can see that the model seems to overestimated some data and underestimated others. This tells us that fitting a linear regression may not be the best way to estimate the true value of the missing data. However, our main goal is to find a reasonable way to fill in the missing data, so we can use all of data points as our input variables. Approximate the missing data with a linear model is good enough for the work we are doing.

Now we have taken care of the missing data, we can start to answer the question that we are interested. Are there any negative relationship between the number of vaccines administered and the number of deaths from COVID. It might be more reasonable to look at new deaths everyday instead of cumulative deaths.

First we want to create a new data frame that records new deaths from the deaths data.
```{r}
date_deaths <- aggregate(deaths$Deaths, by=list(Category=deaths$Date), FUN=sum)
date_deaths$new_death <- diff(c(0, as.matrix(date_deaths$x)))
colnames(date_deaths) <- c("date", "total_deaths", "daily_deaths")

# we only consider dates after 2021-01-12 because that's when we started to have data about the number of vaccinated people
date_deaths <- date_deaths %>% filter(date >= "2021-01-12")

plot(date_deaths$date, date_deaths$daily_deaths, xlab="number of people vacinated", ylab = "number of death cases daily")
```
Lets get rid of columns we don't need.
```{r}
vac_death <- merge(tt_vac[ , c("date", "people_vaccinated")], date_deaths[ , c("date", "daily_deaths")], by="date")
```
We can now start or hypothesis test. Recall that we want to know if there is a negative relationship between the number of people get vaccinated and number of daily deaths. We set our null hypothesis to there is no relationship between the two features and alternative hypothesis to there is a negative relationship between the two features. In other words, we can write them as $H_0:\hat{\beta}=0$ and $H_a:\hat{\beta}<0$ with $\hat{\beta}$ as our estimator for true slope for the two parameters that we are interested.

```{r}
beta = 0
x_samp = vac_death$people_vaccinated
y_samp = vac_death$daily_deaths
n = length(vac_death$people_vaccinated)
x_bar = mean(x_samp)
y_bar = mean(y_samp)

# using the functions earlier, we can find an estimate of the parameters of the linear model
beta_hat <- sxy(x_samp, y_samp)/sxx(x_samp)
alpha_hat <- y_bar - beta_hat * x_bar
y_hat <- function(x){beta_hat*x + alpha_hat}

#plot
plot(x_samp, y_hat(x_samp), type='l', xlim = c(min(x_samp),max(x_samp)), ylim = c(min(y_samp),max(y_samp)))
points(x_samp, y_samp)
```

Know we can find the t statistics and p value with the information we have.
```{r}
sse = syy(y_samp) - beta_hat*sxy(x_samp, y_samp)
s_squared = sse/(n-2)

t_stat = (beta_hat - beta)/sqrt(s_squared/sxx(x_samp))
p_val = pt(t_stat, n-2, lower.tail = TRUE)
p_val
```
We get a p value that is so small that any reasonable significance level are greater than it, so we can easily conclude that we reject our null hypothesis and say that number of people vaccinated has a negative relationship with daily death cases.

Here is the slope of the linear model.
```{r}
beta_hat
1/beta_hat
```
The slope represent the changes in daily death cases for every 1 person get vaccinated. The reciprocal shows the changes in number of people vaccinated for every dead person. In other words, for every 4k Americans get vaccinated, there is one person been saved. This statistics shows that fighting the pandemic is a team work. It requires people to work as a country to brings the numbers down.
